---
title: "Final Project Appendix"
author: "Alex G. and Max M."
date: "5/31/2017"
output: pdf_document
---

```{r}
library(car)
library(lattice)
econ <- read.csv("/Users/alexgriese/College/Math/Applied Regression Analysis/Final Project/finaldata.csv", header = TRUE, sep = ",")

#Creating ratio of imports to exports
econ$TradeRatio <- econ$Imports/econ$Exports
econ
```

##Performing EDA:

```{r}
scatterplotMatrix(~ GDP_Rate + Inflation + IntRate + PrimEducRate +SecEducRate + TerEducRate, data = econ)
scatterplotMatrix(~ GDP_Rate + MCapRate+ Population + TradeRatio + Unemployment + S.P500 + Poverty, data = econ)
```

We see that some of our variables (specifically MCapRate and Population) appear to be highly colinear (we will deal with this later).

A transformations is needed to Inflation. It seems a log transformation aids somewhat in creating a better relationship between GDP_Rate and Inflation. \textit{Note:} in order to have a log transformation, we must insure that all the inflation data points are positive. There is one data point less than 0$-$ -0.36\% to be exact$-$as such we merely add 1 to every variable in the data set.

```{r,fig.height=4}

par(mfrow = c(1,2))
plot(GDP_Rate ~ (Inflation) , data = econ)
plot(GDP_Rate ~ log(I(Inflation+1)) , data = econ)
par(mfrow = c(1,1))
```

A transformations is needed to IntRate It seems a log transformation aids somewhat in creating a better relationship between GDP_Rate and Inflation. \textit{Note:} in order to have a log transformation, we must insure that all the IntRate data points are positive. There is one data point less than 0$-$ -1.28\% to be exact$-$as such we merely add 1.29 to every variable in the data set.

IntRate might have some kind of interaction term (with what?)

```{r,fig.height=4}

par(mfrow = c(1,2))
plot(GDP_Rate ~ (IntRate) , data = econ)
plot(GDP_Rate ~ log(I(Inflation+1.29)) , data = econ)
par(mfrow = c(1,1))
```