---
title: "Final Project Appendix"
author: "Alex G. and Max M."
date: "5/31/2017"
output: pdf_document
---

```{r}
library(car)
library(lattice)
econ <- read.csv("https://raw.githubusercontent.com/griesea/Math245/master/finaldata.csv", header = TRUE, sep = ",")
#Creating ratio of imports to exports
econ$TradeRatio <- econ$Imports/econ$Exports
econ
```

##Performing EDA:

```{r}
scatterplotMatrix(~ GDP_Rate + Inflation + IntRate + PrimEducRate +SecEducRate + TerEducRate, data = econ)
scatterplotMatrix(~ GDP_Rate + MCapRate+ Population + TradeRatio + Unemployment + S.P500 + Poverty, data = econ)
```

We see that some of our variables (specifically MCapRate and Population) appear to be highly colinear (we will deal with this later).

###Transformations

Inflation looks like it needs a transformation. However, economic theory suggests a peak benefit inflation rate (with too low or too high inflation rates being detrimental to the economy). As such, we will try to make inflation a quadratic term in the model, and so do not transform it here.


We dont appear to need a tranformation to IntRate (no convincing trend appears).

```{r,fig.height=4}
plot(GDP_Rate ~ (IntRate) , data = econ)
```


Trade ratio, poverty rate, and unemployment show no convincing needs for transformations:
```{r, fig.height = 4}
par(mfrow=c(1,3))
plot(GDP_Rate ~ (TradeRatio), data=econ)
plot(GDP_Rate ~ Unemployment, data=econ)
plot(GDP_Rate ~ Poverty, data=econ)
par(mfrow=c(1,1))
```

However, S.P500 might have a quadratic effect on GDP_Rate. We will test this as we fit our model. 


###Preliminary Linear Model
```{r}
#test.lm1 <- lm(GDP_Rate ~ )
par(mfrow = c(1,2))
plot(GDP_Rate ~ (S.P500) , data = econ)
#plot(GDP_Rate ~ sqrt(S.P500) , data = econ)
par(mfrow = c(1,1))
```
```

